\bf Model Visualization Techniques for a Social Network Model

### 2017-08-22 paper.Rnw
a couple blurbs of discussion.

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  axis.title.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(2),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes. When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account. 
Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be. A social network does not form spontaneously: it evolves over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

@@

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
Note here that our edge variables are binary: we only consider the presence or absence of an edge. If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

Each type of visualization comes with its own advantages and disadvantages. For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges. Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.  \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} \st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  \hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}. Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$. Then, the probility that the tie $x_{ij}$ changes is

@@

<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to), 
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>% 
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>% 
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1. The occurrence percent is split up into groups to correspond with its distribution: only about 10\% of the edges appear more than 10\% of the time in the 1,000 simulations, while about 60\% appear less than 1\% of the time. The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison. We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
# get a plot of all microsteps from w1 to w2 by edge
alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
alledges$eid <- 1:nrow(alledges)

@@

>>>>

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.title.x = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(1.5),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{%XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes. When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account. 
Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be. A social network does not form spontaneously: it evolves over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

@@

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
Note here that our edge variables are binary: we only consider the presence or absence of an edge. If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

Each type of visualization comes with its own advantages and disadvantages. For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges. Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.  %\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} 
\st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

@@

%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. %\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?} 
Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}. Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$. Then, the probility that the tie $x_{ij}$ changes is

@@

<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to), 
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>% 
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>% 
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1. The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time. The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison. We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
# get a plot of all microsteps from w1 to w2 by edge
alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
alledges$eid <- 1:nrow(alledges)

@@

We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented model for social network. By looking at the underlying algorithms, exploring collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.

We have only just begun to scratch the surface of these complicated and multi-layered models for social networks. The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here. If a researcher thinks the network structure or an actor covariate effects the rate of change of the networ, there is a way to incorporate that belief into the rate function of the SAOM. More than one actor-level covariate can be included in the model, and way more than three parameters can be included in the objective function itself. In addition, \texttt{RSiena} allows the user to tell it which parameters lead to tie creation, and which parameters lead to tie endowment, or dissolution. We have used ``evaluation" parameters, which assume that creation and endowment are equal \citep{RSienamanual}. Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
